\textbf{Selective DNA and Protein Isolation from Marine Macrophyte Surfaces}

### 2020-11-16 submission/manuscript.tex
Coauthors review done.

\usepackage{lineno}

@@

Marino Korlević\textsuperscript{1\(*\)}, Marsej
Markovski\textsuperscript{1}, Zihao Zhao\textsuperscript{2}, Gerhard J.
Herndl\textsuperscript{2,3}, Mirjana Najdek\textsuperscript{1}

@@

@@

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}

@@

Studies of unculturable microbes often combine methods based on DNA,
such as 16S rRNA sequencing and metagenomics, with methods that allow
insight into the metabolic status, such as metaproteomics. To apply
these techniques to the microbial community inhabiting the surfaces of
marine macrophytes it is advisable to perform, prior to the analysis, a
selective DNA and protein isolation so that the host material, present
in higher quantities, is not hampering the analysis. Two protocols, for
DNA and protein isolation, were adapted for selective extractions of DNA
and proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes: the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. Protocols showed an almost
complete removal of the epiphytic community regardless of the sampling
season, station, settlement or host species. The obtained DNA was
suitable for metagenomic and 16S rRNA sequencing, while isolated
proteins could be identified by mass spectrometry, showing that
protocols can be used in 16S rRNA, metagenomic and metaproteomic
analysis. Low presence of host DNA and proteins, observed in isolated
samples, indicated a selective nature of the protocols. Furthermore, the
procedures are based on universally available laboratory chemicals
making the protocols widely applicable. Taken together, the adapted
protocols ensure an almost complete removal of the macrophyte epiphytic
community, are selective for microbes inhabiting macrophyte surfaces and
are providing DNA and proteins applicable in 16S rRNA sequencing,
metagenomics and metaproteomics.

@@

Surfaces of marine macrophytes are inhabited by a diverse microbial
community whose structure and function are poorly understood (Egan
\emph{et al.}, 2013). As less than 1 \% of all prokaryotic species are
culturable, to study these organisms, molecular methods such as 16S rRNA
sequencing, metagenomics and metaproteomics are indispensable (Amann
\emph{et al.}, 1995; Su \emph{et al.}, 2012). Applying these techniques
requires an initial isolation step, with the purpose of obtaining high
quality DNA and proteins.

Biological material (i.e.~proteins and DNA) from pelagic microbial
communities is usually isolated by collecting cells onto filters and
subsequently isolating the aimed compound (Gilbert \emph{et al.}, 2009).
If a specific microbial size fraction is aimed sequential filtration is
applied (Massana \emph{et al.}, 1997; Andersson \emph{et al.}, 2010). In
contrast, obtaining biological materials from microorganism inhabiting
surfaces requires either a cell detachment procedure prior to isolation
or the host material is coextracted together with the targeted material.
Methods for separating microbial cells form the host include shaking of
host tissue (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping of macrophyte surfaces (Uku \emph{et al.}, 2007) or the
application of ultrasonication (Weidner \emph{et al.}, 1996; Cai
\emph{et al.}, 2014). It was shown that shaking alone is not sufficient
to remove microbial cells, at least from plant root surfaces
(Richter-Heitmann \emph{et al.}, 2016). Manual separation methods, such
as scraping and brushing, are time consuming and subjective, as the
detachment efficiency depends on host tissue and the person performing
the procedure (Cai \emph{et al.}, 2014). Ultrasonication was proposed as
an alternative method as it is providing better results in terms of
detachment efficiency (Cai \emph{et al.}, 2014; Richter-Heitmann
\emph{et al.}, 2016). The downside of this procedure is that complete
cell removal was still not obtained and tissue disruption was observed
especially after the application of probe ultrasonication
(Richter-Heitmann \emph{et al.}, 2016). An alternative to these cell
detachment procedures is the isolation of targeted epiphytic compounds
together with host materials (Staufenberger \emph{et al.}, 2008; Jiang
\emph{et al.}, 2015). This procedure can lead to problems in the
following processing steps such as mitochondrial and chloroplast 16S
rRNA sequence contaminations from the host (Longford \emph{et al.},
2007; Staufenberger \emph{et al.}, 2008). In addition, when performing
metagenomics and metaproteomics host material can cause biased results
towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the targeted
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed and the isolation
procedure continues, omitting host material contaminations. To our
knowledge the only procedure describing a direct and selective epiphytic
DNA isolation from the surfaces of marine macrophytes was provided by
Burke \emph{et al.} (2009). In contrast to previously described methods
this protocol enables an almost complete removal of the surface
community and it was used for 16S rRNA gene clone library construction
(Burke and Thomas \emph{et al.}, 2011) and metagenome sequencing (Burke
and Peter Steinberg \emph{et al.}, 2011). This method, though providing
a selective isolation procedure, uses a rapid multienzyme cleaner (3M)
that is not available worldwide and without a known composition (Burke
\emph{et al.}, 2009). Also to our knowledge, no selective isolation
protocol for proteins from epiphytic communities inhabiting marine
macrophytes was established.

In the present study, we adapted a protocol widely used for DNA
isolation from filters (Massana \emph{et al.}, 1997) and a protocol used
for protein isolation from sediments (Chourey \emph{et al.}, 2010;
Hultman \emph{et al.}, 2015) for selective extractions of DNA and
proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes: the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. In addition, we tested the
removal efficiency of the protocol and the suitability of obtained DNA
and proteins for 16S rRNA sequencing, metagenomics and metaproteomics.

@@

To assess the removal efficiency of DNA and protein isolation procedures
leaves and thalli were examined under a confocal microscope before and
after treatments were performed. Developed procedures showed an almost
complete removal of the surface community of both, \emph{C. nodosa} and
\emph{C. cylindracea}. In addition, a similar removal efficiency was
observed for communities sampled in contrasting seasons, December 2017
(\autoref{confocal_december}) and June 2018 (\autoref{confocal_june}).
Also, no effect of station, settlement or isolation procedure (DNA or
protein) on the removal efficiency was observed (Figs.
\ref{confocal_december} and \ref{confocal_june}).

To evaluate if the obtained DNA is suitable to determine the microbial
community structure an Illumina sequencing of the V4 16S rRNA region was
performed. Sequencing yielded a total of 336,944 sequences after quality
curation and exclusion of eukaryotic, mitochondrial and no relative
sequences. The number of sequences classified as chloroplast was 97,334.
After excluding these sequences the total number of retrieved reads
dropped to 239,610, ranging from 22,596 to 52,930 sequences per sample
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off which is commonly
observed in high-throughput 16S rRNA amplicon sequencing procedures
(\autoref{supp-rarefaction}). Sequence clustering at a similarity level
of 97 \si{\percent} yielded a total of 8,355 different OTUs. Taxonomic
classification of reads allowed for a macrophyte associated epiphytic
community determination that was mainly composed of:
\emph{Alphaproteobacteria} (14.9 ± 3.5 \si{\percent}),
\emph{Bacteroidota} (12.5 ± 2.4 \si{\percent}),
\emph{Gammaproteobacteria} (11.6 ± 4.3 \si{\percent}),
\emph{Desulfobacterota} (7.8 ± 7.5 \si{\percent}), \emph{Cyanobacteria}
(6.5 ± 4.7 \si{\percent}) and \emph{Planctomycetota} (2.9 ± 1.7
\si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon
procedures, also amplified chloroplast 16S rRNA genes. We observed a
high proportion of chloroplast sequences in all analyzed samples (33.4 ±
9.4 \si{\percent}) (\autoref{community}). To determine if chloroplast
sequences originate from hosts or eukaryotic epiphytic organisms, we
exported SILVA classified chloroplast sequences and reclassified them
using the RDP (Ribosomal Database Project) training set that allows for
a more detailed chloroplast classification. The largest proportion of
sequences were classified as Bacillariophyta (89.7 ± 5.7 \si{\percent})
indicating that the DNA removal procedure did not coextract larger
quantities of host DNA (\autoref{chloroplast}). Chloroplast sequences
classified as Streptophyta constituted 3.3 ± 2.8 \si{\percent} of all
chloroplast sequences originating from \emph{C. nodosa} samples, while
sequences classified as Chlorophyta comprised only 0.02 ± 0.01
\si{\percent} of all chloroplast sequences associated with \emph{C.
cylindracea} samples.

To determine if the extracted DNA can be used for metagenomic sequencing
four samples containing epiphytic DNA were selected and shotgun
sequenced using an Illumina platform. Metagenomic sequencing yielded
from 207,149,524 to 624,029,930 sequence pairs
(\autoref{supp-metagenomic_statistics}). Obtained sequences were
successfully assembled into contigs whose L50 ranged from 654 to 1,011
bp. In addition, predicted coding sequences were successfully
functionally annotated (9,066,667 -- 20,256,215 annotated sequences;
\autoref{cog}) and taxonomically classified. Functional annotation
allowed for an assessment of the relative contribution of each COG
(Clusters of Orthologous Groups) functional category to the total number
of annotated coding sequences (\autoref{cog}a). Functional categories
containing the highest number of sequences were: C (Energy production
and conversion), E (Amino acid transport and metabolism), M (Cell
wall/membrane/envelope biogenesis), L (Replication, recombination and
repair) and P (Inorganic ion transport and metabolism). If host DNA was
coextracted with epiphytic it should be detected in larger proportions
in sequenced metagenomes. Indeed, no higher proportions of coding
sequences classified into phylum Streptophyta or Chlorophyta were
detected (\autoref{supp-metagenomic_taxonomy}). Sequenced metagenomic
DNA originating from the surface of \emph{C. nodosa} contained 1.3
\si{\percent} of coding sequences classified into the phylum
Streptophyta in December 2017 and 0.7 \si{\percent} in June 2018.
Furthermore, the summed RPKM (Reads Per Kilobase Million) of these
sequences constituted 1.7 \si{\percent} of total RPKM of all
successfully classified sequences in December 2017 and 1.1 \si{\percent}
in June 2018. Similar low proportions of host coding sequences were
detected in metagenomic samples originating from the surfaces of
\emph{C. cylindracea}. Of all successfully classified coding sequences
0.2 \si{\percent} were classified into Chlorophyta in December 2017 and
0.1 \si{\percent} in June 2018. A relatively higher proportion of these
sequences' RPKM than in the case of \emph{C. nodosa} was observed,
indicating a higher coextraction of host DNA in \emph{C. cylindracea}.
In December, the proportion of RPKM of sequences classified into
Chlorophyta was 8.2 \si{\percent}, while in June 2018 it reached 13.6
\si{\percent}.

To evaluate if the procedure for protein extraction is suitable for
metaproteomic analysis, obtained proteins were trypsin digested and
sequenced using a mass spectrometer. Obtained MS/MS spectra were
searched against a protein database from sequenced metagenomes. From
14,219 to 16,449 proteins were identified in isolated protein samples
(\autoref{cog}b). In addition, successful identification of proteins
allowed for an assessment of the relative contribution of each COG
functional category to the total number of identified proteins
(\autoref{cog}b). Functional categories containing the highest number of
identified proteins were: C (Energy production and conversion), G
(Carbohydrate transport and metabolism), P (Inorganic ion transport and
metabolism), O (Posttranslational modification, protein turnover,
chaperones) and E (Amino acid transport and metabolism). Isolated
proteins could derive from epiphytic organisms inhabiting the macrophyte
surface or from macrophyte tissue underlying them. The contribution of
proteins originating from host tissues was evaluated by identifying all
the proteins predicted to belong to any taxonomic group within the phyla
Streptophyta and Chlorophyta and by calculating their contribution to
the number and abundance of all identified proteins. On average,
proteins isolated from the surface of \emph{C. nodosa} contained 1.8 ±
0.06 \si{\percent} of proteins associated with Streptophyta,
contributing to 2.2 ± 0.8 \si{\percent} of total proteins. Similar to
metagenomes, proteins associated with Chlorophyta contributed more to
\emph{C. cylindracea} than proteins associated with Streptophyta to
\emph{C. nodosa}. Chlorophyta associated proteins comprised 5.2 ± 0.06
\si{\percent} of all identified proteins in \emph{C. cylindracea},
contributing to 19.2 ± 1.5 \si{\percent} of all protein abundances.

@@

The study of marine macrophyte epiphytic communities using culture
independent techniques, such as 16S rRNA analysis, metagenomics and
metaproteomics, requires an initial step of biological material
isolation. Methods that have been developed for selective isolation of
epiphytic biological material can be divided into two groups: (i)
procedures involving, prior to extraction, a cell detachment step such
as shaking (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping (Uku \emph{et al.}, 2007) and ultrasonication (Weidner \emph{et
al.}, 1996; Cai \emph{et al.}, 2014) and (ii) procedures involving a
host tissue incubation aiming at direct lysis of epiphytic microbial
cells (Burke \emph{et al.}, 2009). Protocols that include a cell
detachment procedure do not provide complete cell removal (Cai \emph{et
al.}, 2014; Richter-Heitmann \emph{et al.}, 2016) and in the case of
probe ultrasonication tissue disruption can also be observed
(Richter-Heitmann \emph{et al.}, 2016). Though, host tissue incubation
procedures provide an almost complete cell removal and selective
isolation, existing protocols like the one developed for DNA isolation
by Burke \emph{et al.} (2009) use in the extraction buffer a rapid
multienzyme cleaner (3M) not available worldwide and without a known
composition (Burke \emph{et al.}, 2009). To circumvent these problems we
developed and tested two protocols for selective DNA and protein
isolation from marine macrophyte epiphytic communities.

To test if the developed DNA and protein isolation procedures could be
applied on a variety of macrophyte species we selected \emph{C. nodosa}
and \emph{C. cylindracea} as representatives of seagrass and macroalgal
species, on which the procedures were tested. These species especially
differ morphologically. While \emph{C. nodosa} leaves are flat, \emph{C.
cylindracea} thallus is composed of uneven surfaces (Kuo and den Hartog,
2001; Verlaque \emph{et al.}, 2003). The developed procedures showed an
almost complete removal of epiphytic cells from the surfaces of both
species comparable to the result of Burke \emph{et al.} (2009) and
indicating that structural differences do not impact the removal
efficiency. In addition, isolation protocols were tested in two
contrasting season as it is known that macrophytes are harboring more
algal epiphytes during autumn and winter (Reyes and Sansón, 2001). No
differences in the removal efficiency was observed between seasons
suggesting that protocols can be used on macrophyte samples retrieved
throughout the year. Also, no removal differences were observed on
samples derived from the same host but from different localities or
settlements demonstrating the stability of the protocol in cell removal
efficiency.

Successful amplification and sequencing of the V4 16S rRNA gene region
proved that the isolated DNA can be used to estimate the microbial
epiphytic diversity. Taxonomic groups detected in this step can also be
often found in epiphytic communities associated with other macrophytes
(Burke and Thomas \emph{et al.}, 2011; Morrissey \emph{et al.}, 2019). A
problem often encountered in studies focusing on epiphytic communities
is the presence of large proportions of chloroplast 16S rRNA sequences
in the pool of amplified molecules, especially if the epiphytic DNA was
isolated without prior selection (Staufenberger \emph{et al.}, 2008).
These sequences can derive from host chloroplasts or from eukaryotic
epiphytic chloroplast DNA. Although the proportion of obtained
chloroplast 16S rRNA sequences in our samples was substantial they
derived almost exclusively from eukaryotic epiphytes. High proportion of
chloroplast 16S rRNA sequences in studies applying selective procedures
that include direct cellular lysis on host surfaces were observed before
(Michelou \emph{et al.}, 2013). It is possible that chloroplast specific
sequences even in these studies are originating from eukaryotic
epiphytic cells and not host chloroplast. Indeed, it is common during
16S rRNA profiling of pelagic microbial communities to observe high
proportions of chloroplast sequences (Gilbert \emph{et al.}, 2009;
Korlević \emph{et al.}, 2016). In addition, a very low proportion of
chloroplast 16S rRNA sequences in samples originating from \emph{C.
cylindracea} in comparison to \emph{C. nodosa} could be explained by the
presence of three introns in the gene for 16S rRNA in some members of
the genus \emph{Caulerpa} that could hamper the amplification process
(Lam and Lopez-Bautista, 2016).

Beside 16S rRNA sequencing high quality DNA is needed for metagenomics.
The obtained number of metagenomic sequences and assembly statistics
were comparable to metagenomes and metatranscriptomes derived from
similar surface associated communities (Crump \emph{et al.}, 2018; Cúcio
\emph{et al.}, 2018). In addition, functional annotation of predicted
coding sequences to COG functional categories showed that the obtained
metagenomes can be used to determine the metabolic capacity of surface
associated communities (Leary \emph{et al.}, 2014; Cúcio \emph{et al.},
2018). The proportion of coding sequences, including their RPKM,
originating from \emph{C. nodosa} metagenomes and classified as
Streptophyta was low indicating a good selectivity of the isolation
procedure towards epiphytic cells. In the case of DNA samples isolated
from the surface of \emph{C. cylindracea} the proportion of Chlorophyta
coding sequences was also low but their RPKM was higher than in the case
of \emph{C. nodosa}. One of the causes for this elevated RPKM of
Chlorophyta sequences in \emph{C. cylindracea} could lay in tissue
structure differences between these two host species. While \emph{C.
nodosa} leaves are composed of individual cells, the thallus of \emph{C.
cylindracea} is, like in other siphonous algal species, composed of a
single large multinucleate cell (Coneva and Chitwood, 2015). The absence
of individual cells in \emph{C. cylindracea} could cause a leakage of
genetic material into the extraction buffer causing an elevated presence
of host sequences in metagenomic data.

>>>>

\usepackage{lineno}
\renewcommand{\linenumberfont}{\normalfont\tiny}

@@

Marino Korlević\textsuperscript{1\(*\)}, Marsej
Markovski\textsuperscript{1}, Zihao Zhao\textsuperscript{2}, Gerhard J.
Herndl\textsuperscript{2,3,4}, Mirjana Najdek\textsuperscript{1}

@@

4. Vienna Metabolomics Center, University of Vienna, Austria

@@

\sisetup{mode=text}
\setlength\parindent{24pt}

@@

Studies of unculturable microbes often combine methods based on DNA,
such as 16S rRNA sequencing and metagenomics, with methods that allow
insights into the metabolic status, such as metaproteomics. To apply
these techniques to the microbial community inhabiting the surfaces of
marine macrophytes it is advisable to perform a selective DNA and
protein isolation prior to the analysis to avoid biases due to the host
material being present in high quantities. Two protocols for DNA and
protein isolation were adapted for selective extractions of DNA and
proteins from epiphytic communities inhabiting the surfaces of two
marine macrophytes, the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea}. Protocols showed an almost
complete removal of the epiphytic community regardless of the sampling
season, station, settlement or host species. The obtained DNA was
suitable for metagenomic and 16S rRNA sequencing, while isolated
proteins could be identified by mass spectrometry. The low presence of
host DNA and proteins in the samples indicated a high specificity of the
protocols. The procedures are based on universally available laboratory
chemicals making the protocols widely applicable. Taken together, the
adapted protocols ensure an almost complete removal of the macrophyte
epiphytic community. The procedures are selective for microbes
inhabiting macrophyte surfaces and provide DNA and proteins applicable
in 16S rRNA sequencing, metagenomics and metaproteomics.

@@

Surfaces of marine macrophytes are colonized by a diverse microbial
community whose structure and function are poorly understood (Egan
\emph{et al.}, 2013). As less than 1 \% of all prokaryotic species are
culturable, molecular methods such as 16S rRNA sequencing, metagenomics
and metaproteomics are indispensable to study these organisms (Amann
\emph{et al.}, 1995; Su \emph{et al.}, 2012). Applying these techniques
requires an initial isolation step with the purpose of obtaining high
quality DNA and proteins.

Biological material (i.e., proteins and DNA) from pelagic microbial
communities is usually isolated by collecting cells onto filters and
subsequently isolating the target organisms or communities (Gilbert
\emph{et al.}, 2009). If a specific microbial size fraction is aimed
sequential filtration is applied (Massana \emph{et al.}, 1997; Andersson
\emph{et al.}, 2010). In contrast, obtaining microorganisms associated
to surfaces require either a cell detachment procedure prior to
isolation or the host material is co-extracted with the target material.
Methods for separating microbial cells from the host include shaking of
host tissue (Gross \emph{et al.}, 2003; Nõges \emph{et al.}, 2010),
scraping of macrophyte surfaces (Uku \emph{et al.}, 2007) or applying
ultrasonication (Weidner \emph{et al.}, 1996; Cai \emph{et al.}, 2014).
It was shown that shaking alone is not sufficient to remove microbial
cells from surfaces, at least not from plant root surfaces
(Richter-Heitmann \emph{et al.}, 2016). Manual separation methods, such
as scraping and brushing are time consuming and subjective, as the
detachment efficiency depends on host tissue and the person performing
the procedure (Cai \emph{et al.}, 2014). Ultrasonication was proposed as
an alternative method as it is providing better results in terms of
detachment efficiency (Cai \emph{et al.}, 2014; Richter-Heitmann
\emph{et al.}, 2016). The downside of this procedure is that complete
cell removal is still not obtained and tissue disruption was observed
especially after the application of probe ultrasonication
(Richter-Heitmann \emph{et al.}, 2016). An alternative to these cell
detachment procedures is the isolation of target epiphytic compounds
together with host material (Staufenberger \emph{et al.}, 2008; Jiang
\emph{et al.}, 2015). This procedure can lead to problems in the
following processing steps such as mitochondrial and chloroplast 16S
rRNA sequence contaminations from the host (Longford \emph{et al.},
2007; Staufenberger \emph{et al.}, 2008). In addition, when performing
metagenomics and metaproteomics host material can cause biased results
towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the target
material by incubating macrophyte tissues in an extraction buffer. After
the incubation, the undisrupted host tissue is removed followed by the
isolation procedure, omitting host material contaminations. To our
knowledge, the only procedure describing a direct and selective
epiphytic DNA isolation from the surfaces of marine macrophytes was
described by Burke \emph{et al.} (2009). In contrast to previously
described methods, this protocol enables an almost complete removal of
the surface community. It was used for 16S rRNA gene clone library
construction (Burke and Thomas \emph{et al.}, 2011) and metagenome
sequencing (Burke and Peter Steinberg \emph{et al.}, 2011). This method,
although providing a selective isolation procedure, uses a rapid
multi-enzyme cleaner (3M) that is not available worldwide and the
chemical constituents are unknown (Burke \emph{et al.}, 2009). Also to
our knowledge, no selective isolation protocol to perform
(meta)proteomics of epiphytic communities associated with marine
macrophytes has been developed yet.

In the present study, we adapted a protocol widely used for DNA
isolation from filters (Massana \emph{et al.}, 1997) and a protocol used
for protein isolation from soils (Chourey \emph{et al.}, 2010; Hultman
\emph{et al.}, 2015). These two adapted methods allowed for a selective
extraction of DNA and proteins from epiphytic communities inhabiting the
surfaces of two marine macrophytes, the seagrass \emph{Cymodocea nodosa}
and the macroalga \emph{Caulerpa cylindracea}. In addition, we tested
the removal efficiency of the protocol and the suitability of obtained
DNA and proteins for 16S rRNA sequencing, metagenomics and
metaproteomics.

@@

To assess the removal efficiency of the DNA and protein isolation
procedures, leaves and thalli were examined under a confocal microscope
before and after treatments were performed. The modified procedures
resulted in an almost complete removal of the surface community of both,
\emph{C. nodosa} and \emph{C. cylindracea}. In addition, a similar
removal efficiency was observed for communities sampled in contrasting
months, December 2017 (\autoref{confocal_december}) and June 2018
(\autoref{confocal_june}). Also, no effect of station, settlement or
isolation procedure (DNA or protein) on the removal efficiency was
observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

To evaluate whether the obtained DNA is suitable to determine the
composition of the microbial community Illumina sequencing of the V4
region of the 16S rRNA was performed. Sequencing yielded a total of
336,944 sequences after quality curation and exclusion of eukaryotic,
mitochondrial and no relative sequences. The number of sequences
classified as chloroplasts was 97,334. After excluding these sequences
the total number of retrieved reads was 239,610, ranging from 22,596 to
52,930 sequences per sample (\autoref{supp-nseq_notus}). Even when the
highest sequencing effort was applied, the rarefaction curves did not
level off which is commonly observed in high-throughput 16S rRNA
amplicon sequencing (\autoref{supp-rarefaction}). Sequences clustering
at a similarity level of 97 \si{\percent} yielded a total of 8,355
different OTUs. Taxonomic classification of reads revealed a
macrophyte-associated epiphytic community that mainly composed of
\emph{Alphaproteobacteria} (14.9 ± 3.5 \si{\percent}),
\emph{Bacteroidota} (12.5 ± 2.4 \si{\percent}),
\emph{Gammaproteobacteria} (11.6 ± 4.3 \si{\percent}),
\emph{Desulfobacterota} (7.8 ± 7.5 \si{\percent}), \emph{Cyanobacteria}
(6.5 ± 4.7 \si{\percent}) and \emph{Planctomycetota} (2.9 ± 1.7
\si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon
procedures, also amplified chloroplast 16S rRNA genes. We observed a
high proportion of chloroplast sequences in all analyzed samples (33.4 ±
9.4 \si{\percent}) (\autoref{community}). To determine whether
chloroplast sequences originate from the host or eukaryotic epiphytic
organisms, we exported SILVA-classified chloroplast sequences and
reclassified them using the RDP (Ribosomal Database Project) training
set that allows for a more detailed chloroplast classification. The
largest proportion of sequences was classified as Bacillariophyta (89.7
± 5.7 \si{\percent}) indicating that the DNA removal procedure resulted
in only minor co-extracted quantities of host DNA
(\autoref{chloroplast}). Chloroplast sequences classified as
Streptophyta constituted 3.3 ± 2.8 \si{\percent} of all chloroplast
sequences originating from \emph{C. nodosa} samples, while sequences
classified as Chlorophyta comprised only 0.02 ± 0.01 \si{\percent} of
all chloroplast sequences associated with \emph{C. cylindracea} samples.

To determine whether the extracted DNA can be used for metagenomic
sequencing, four samples containing epiphytic DNA were selected and
shotgun sequenced using an Illumina platform. Metagenomic sequencing
yielded between 207,149,524 and 624,029,930 sequence pairs
(\autoref{supp-metagenomic_statistics}). Obtained sequences were
successfully assembled into contigs whose L50 ranged from 654 to 1,011
bp. In addition, predicted coding sequences were functionally annotated
(9,066,667 -- 20,256,215 annotated sequences; \autoref{cog}) and
taxonomically classified. Functional annotation allowed for an
assessment of the relative contribution of each COG (Clusters of
Orthologous Groups) functional category to the total number of annotated
coding sequences (\autoref{cog}a). Functional categories containing the
highest number of sequences were C (Energy production and conversion), E
(Amino acid transport and metabolism), M (Cell wall/membrane/envelope
biogenesis), L (Replication, recombination and repair) and P (Inorganic
ion transport and metabolism). If host DNA is co-extracted with
epiphytes it should be detected in large proportions in sequenced
metagenomes. However, no large proportions of coding sequences
classified as Streptophyta and Chlorophyta were detected
(\autoref{supp-metagenomic_taxonomy}). Sequenced metagenomic DNA
originating from the surface of \emph{C. nodosa} contained 1.3
\si{\percent} of coding sequences classified as Streptophyta in December
2017 and 0.7 \si{\percent} in June 2018. Furthermore, the summed RPKM
(Reads Per Kilobase Million) of these sequences constituted 1.7
\si{\percent} of total RPKM of all successfully classified sequences in
December 2017 and 1.1 \si{\percent} in June 2018. Similar low
proportions of host coding sequences were detected in metagenomic
samples originating from the surfaces of \emph{C. cylindracea}. Of all
successfully classified coding sequences 0.2 \si{\percent} were
classified as Chlorophyta in December 2017 and 0.1 \si{\percent} in June
2018. A relatively higher proportion of RPKM of these sequences than in
the case of \emph{C. nodosa} was observed, indicating a higher
coextraction of host DNA in \emph{C. cylindracea}. In December, the
proportion of RPKM of sequences classified as Chlorophyta was 8.2
\si{\percent}, while in June 2018 it reached 13.6 \si{\percent}.

To evaluate whether the procedure for protein extraction is suitable for
metaproteomic analysis, obtained proteins were trypsin digested and
sequenced using a mass spectrometer. Obtained MS/MS spectra were
searched against a protein database from sequenced metagenomes. From
14,219 to 16,449 proteins were identified in isolated protein samples
(\autoref{cog}b). In addition, successful identification of proteins
allowed for an assessment of the relative contribution of each COG
functional category to the total number of identified proteins
(\autoref{cog}b). Functional categories containing the highest number of
identified proteins were C (Energy production and conversion), G
(Carbohydrate transport and metabolism), P (Inorganic ion transport and
metabolism), O (Posttranslational modification, protein turnover,
chaperones) and E (Amino acid transport and metabolism). Isolated
proteins could originate from epiphytic organisms inhabiting the
macrophyte surface and/or from macrophyte tissue underlying them. The
contribution of proteins originating from host tissues was evaluated by
identifying all the proteins predicted to belong to any taxonomic group
within the phyla Streptophyta and Chlorophyta and by calculating their
contribution to the number and abundance (NAAF -- Normalized Abundance
Area Factor) of all identified proteins. On average, proteins isolated
from the surface of \emph{C. nodosa} contained 1.8 ± 0.06 \si{\percent}
of proteins associated with Streptophyta, contributing to 2.2 ± 0.8
\si{\percent} of total proteins. Similar to metagenomes, proteins
associated with Chlorophyta contributed more to \emph{C. cylindracea}
than proteins associated with Streptophyta to \emph{C. nodosa}.
Chlorophyta associated proteins comprised 5.2 ± 0.06 \si{\percent} of
all identified proteins in \emph{C. cylindracea}, contributing 19.2 ±
1.5 \si{\percent} to the total protein abundance.

@@

To test whether the developed DNA and protein isolation protocols
efficiently detach microbes from the macrophyte surface we selected
\emph{C. nodosa} and \emph{C. cylindracea} as representatives of
seagrass and macroalgal species. These species differ morphologically.
While \emph{C. nodosa} leaves are flat, \emph{C. cylindracea} thalli are
characterized by an uneven surface (Kuo and den Hartog, 2001; Verlaque
\emph{et al.}, 2003). The developed protocol led to an almost complete
removal of epiphytic cells from the surfaces of both species comparable
to the result of Burke \emph{et al.} (2009), indicating that structural
differences do not impact the removal efficiency. In addition, isolation
protocols were tested in two contrasting seasons, as it is known that
macrophytes are harboring more algal epiphytes during autumn and winter
(Reyes and Sansón, 2001). No differences in the removal efficiency was
observed between seasons suggesting that these protocols can be used on
macrophyte samples retrieved throughout the year. Also, no removal
differences were observed on samples derived from the same host but from
different locations.

Successful amplification and sequencing of the V4 region of the 16S rRNA
gene proved that the isolated DNA can be used to estimate the microbial
epiphytic diversity. Taxonomic groups detected in this step can also be
often found in epiphytic communities associated with other macrophytes
(Burke and Thomas \emph{et al.}, 2011; Morrissey \emph{et al.}, 2019). A
problem often encountered in studies focusing on epiphytic communities
is the presence of large proportions of chloroplast 16S rRNA sequences
in the pool of amplified molecules, especially if the epiphytic DNA was
isolated without prior selection (Staufenberger \emph{et al.}, 2008).
These sequences can derive from host chloroplasts or from eukaryotic
epiphytic chloroplast DNA. Although the proportion of obtained
chloroplast 16S rRNA sequences in our samples was substantial, they
derived almost exclusively from eukaryotic epiphytes. High proportion of
chloroplast 16S rRNA sequences in studies applying selective procedures
that include direct cellular lysis on host surfaces were observed before
(Michelou \emph{et al.}, 2013). It is possible that chloroplast-specific
sequences even in these studies originated from eukaryotic epiphytic
cells and not from host chloroplasts. Indeed, it is common during 16S
rRNA profiling of pelagic microbial communities to observe high
proportions of chloroplast sequences (Gilbert \emph{et al.}, 2009;
Korlević \emph{et al.}, 2016). In addition, a very low proportion of
chloroplast 16S rRNA sequences in samples originating from \emph{C.
cylindracea} in comparison to \emph{C. nodosa} could be explained by the
presence of three introns in the gene for 16S rRNA in some members of
the genus \emph{Caulerpa} that could hamper the amplification process
(Lam and Lopez-Bautista, 2016).

High quality DNA is also needed for metagenomics. The obtained number of
metagenomic sequences and assembly statistics were comparable to
metagenomes and metatranscriptomes derived from similar surface
associated communities (Crump \emph{et al.}, 2018; Cúcio \emph{et al.},
2018). In addition, functional annotation of predicted coding sequences
to COG functional categories showed that the obtained metagenomes can be
used to determine the metabolic capacity of surface associated
communities (Leary \emph{et al.}, 2014; Cúcio \emph{et al.}, 2018). The
proportion of coding sequences, including their RPKM, originating from
\emph{C. nodosa} metagenomes and classified as Streptophyta was low
indicating that the isolation procedure was specific for epiphytic
cells. DNA samples isolated from the surface of \emph{C. cylindracea}
exhibited a low proportion of Chlorophyta coding sequences, however,
their RPKM was higher than in the samples originating from \emph{C.
nodosa}. One of the reasons for this elevated RPKM of Chlorophyta
sequences in \emph{C. cylindracea} could be the differences in the
tissue structure between these two host species. While \emph{C. nodosa}
leaves are composed of individual cells, the thallus of \emph{C.
cylindracea} is, like in other siphonous algal species, composed of a
single large multinucleate cell (Coneva and Chitwood, 2015). The absence
of individual cells in \emph{C. cylindracea} could cause a leakage of
genetic material into the extraction buffer causing an elevated presence
of host sequences in the samples for metagenome analyses.

To obtain insight into the metabolic status of uncultivated prokaryotes,
a metaproteomic approach is required (Saito \emph{et al.}, 2019). The
applied protocol for epiphytic protein isolation followed by a
metaproteomic analysis identified between 14,219 and 16,449 proteins,
which is higher than previously reported for e.g.~soils (Chourey
\emph{et al.}, 2010; Hultman \emph{et al.}, 2015), seawater (Williams
\emph{et al.}, 2012) and biofilms (Leary \emph{et al.}, 2014). The
functional annotation of identified proteins into COG functional
categories showed that the protein isolation protocol can be used to
assess the metabolic status of the epiphytic community (Leary \emph{et
al.}, 2014). Similar to the results of the metagenomic analysis, the
number and abundance of identified proteins affiliated to Streptophyta
in \emph{C. nodosa} samples were low, indicating that the procedure is
selective for epiphytic cell proteins. In addition, a higher number and
abundance of identified proteins associated with Chlorophyta were
observed in \emph{C. cylindracea} samples. The cause of this elevated
presence of Chlorophyta-associated proteins can be, similar to the DNA
isolation protocol, explained by the absence of individual cells in this
siphonous alga (Coneva and Chitwood, 2015).
